Does It Scale? Static Output Approximation of PHP Web Applications Using Symbolic Execution

### 2017-01-13 paper.tex
refactored introduction

One common property of all technologies for dynamic web applications is
\emph{staged computation} of output: A dynamic web application as a whole
consists of both static code, such as scripts, and dynamically generated code,
such as responses to HTTP requests. The latter code, although assembled at
runtime, may contain client-side parts of the web application, such as
JavaScript. So, to study dynamic web applications in its entirety, we need to
consider both static as well as dynamic aspects of systems.

%Since based on both static values and runtime values, output code is generated step by step
%dynamically, output can be dependent on application inputs. Dynamic web
%applications incorporate various both server-side inputs (databases,
%configuration) and client-side inputs respectively. Static
%analysis of dynamic web applications, hence, has to consider different input
%sources.

\emph{Symbolic execution} is a static analysis technique between program testing
and program proving. For a program, symbolic, i.e., abstract, yet fixed inputs are
applied instead of concrete inputs \cite{Darringer1978,King1976}. The underlying
concept is to map program input (symbolically) to program output. Different classes of inputs may
result in variational output, representing dependencies of output and input.
As symbolic execution of a program returns an input-output-mapping for symbolic
inputs, this analysis technique helps unfold the \emph{staged} nature of web
applications: Since every feasible path is executed all possible output is
contained in the symbolic output, and all dynamically generated variational
parts of the web application can be analyzed.

Knowledge about different output variants enabled useful analyses
for the domain of web applications, such as detecting and locating HTML
validation errors \cite{Nguyen:2011:AFH:2190078.2190142}. As for developers,
from an output model, outlines and callgraphs for easier IDE code navigation
\cite{Nguyen:2014:BCG:2635868.2635928}, or program slices
\cite{Nguyen:2015:CPS:2786805.2786872} can be computed. Existing work has been
integrated in the Varis plugin for the Eclipse IDE \cite{Nguyen:2015:VIS:2819009.2819140}.

Despite the various use cases for analyses based on an approximated output
model, static output approximation for PHP web applications using symbolic
execution has so far only been evaluated for smaller systems that are not
maintained anymore. The previously mentioned tools though are only practical, if
for a given system the symbolic execution engine is scalable, i.e., will also
approximate output accurately for larger and more recent systems with good time
and space consumption.

To investigate the question, whether we can build practical tools from the
the symbolic execution engine, we re-implemented the engine with the specifications
of the previous symbolic execution semantics
\cite{Nguyen:2014:BCG:2635868.2635928} for PHP and additional features, such
as object-oriented programming. We evaluated our
symbolic execution engine for large-scale and modern PHP systems.

During introduction of new semantics for object-orientation we encountered two
main challenges. First, we were bound to introduce more concrete execution features to the
symbolic execution engine, e.g. foreach loops for arrays, as this increased
code coverage significantly. Second, we had to make a trade-off between the
accuracy of method calls and the scalability of this language feature as a
method call for different program state variants can have multiple targets.

Based on our observations and evaluation results, we learned that for dynamic
PHP constructs, expressions that are assembled dynamically, in combination with
symbolic values symbolic execution can be impractical. For object-orientation
and method calls, symbolic execution semantics can be a tightrope walk between
state space explosion and imprecision. Moreover, the nature of PHP challenges a
practical symbolic execution engine since for it to be accurate, a large number
of functions from the standard library needs to be supported.

Our key contributions in this paper include (1) a new tool infrastructure to
statically approximate the output for PHP web applications using symbolic
execution, and (2) a report of our observations with state-of-the-art web
applications as well as an explanation on conceptual limitations for symbolic
execution of PHP applications.

>>>>

One common property of all technologies for dynamic web applications is
\emph{staged computation}: A dynamic web application as a whole
consists of both static code, such as scripts, and dynamically generated code,
such as client page output. The latter code, although assembled at
runtime, may contain client-side parts of the web application, such as
JavaScript. So, to study dynamic web applications in its entirety, we need to
consider both static as well as dynamic aspects of systems.

\emph{Symbolic execution} is a static analysis technique between program testing
and program proving. For a program, symbolic values are
used instead of concrete inputs \cite{Darringer1978,King1976}. The underlying
concept is to map program input to program output: Symbolic values are used and
propagated throughout the symbolic execution and keep dependencies for program
output traceable. This analysis technique helps unfold the \emph{staged} nature
of dynamic web applications. Since every feasible path can be executed the
symbolic output contains both invariant output as well as output variants that
depend on program input.

Knowledge about different output output variants is leveraged by a number of
analyses for the domain of web applications. Previous work presented tools
for detecting and locating HTML validation errors
\cite{Nguyen:2011:AFH:2190078.2190142}, computing program slices across
server-side and client-page code \cite{Nguyen:2015:CPS:2786805.2786872}, or
easing development and maintenance by extending IDE support for web applications
with code navigation
\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928}.

Despite the various use cases for analyses based on an approximated output
model, static output approximation for PHP web applications using symbolic
execution has so far only been evaluated for small systems that are not
maintained anymore. The tools presented by previous work
\cite{Nguyen:2015:VIS:2819009.2819140,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2011:AFH:2190078.2190142}
though are only practical, if for a given system the symbolic execution engine
is scalable, i.e., will also approximate output accurately for larger and more
recent systems with acceptable time and space consumption.

To investigate the question, whether we can have a practical and scalable
symbolic executoin engine, we re-implemented the engine with the specifications
of the previous symbolic execution semantics \cite{Nguyen:2014:BCG:2635868.2635928} for PHP and additional features, such
as object-oriented programming. We evaluate our
symbolic execution engine for large-scale and modern PHP systems.

During the introduction of new semantics, we addressed two  trade-offs between
accuracy of our e xecution and performance: For method calls with ambiguous
targets as well as concrete execution of loops the number of program states can
become infeasible. This may require additional engineering adjustments to tame
the state space explosion at cost of accuracy.

Based on out observations we identify conceptual limitations of symbolic
execution for PHP.  Dynamic features such as indirect function calls and
include expression require concrete information to be evaluated properly. Since
these expressions may be assembled at runtime and can contain symbolic
information, for many cases this restricts our engine from further execution.
Although some effort to statically approximate include expression can be spent,
it is often non-trivial since expressions can contain information coming from
various inputs including databases, user inputs or configuration files. We
empirically evaluate our observations for a corpus including large PHP systems that are still maintained.

Our key contribution in this paper include (1)  a new tool infrastructure to
statically approximate the client-page output for PHP web applications using
symbolic execution and (2) a report of out observations for state-of-the-art
web applications as well as an empirical evaluation of conceptual limitations
of symbolic executions for PHP web applications.
